---
title: "Genome Scans"
author: Owen Kanter
date: "2026-04-23"
---

#Loading Libraries

```{r}
#install.packages("pcadapt")
#install.packages("patchwork")
#install.packages("viridis")
#install.packages("vcfR")
#install.packages("adegenet")
library(pcadapt)
library(vcfR)
library(adegenet)
library(ggplot2)
library(dplyr)
library(readr)
library(patchwork)
library(viridis)
```

# Load Data

```{r}
meta <- read_csv("/home/f85j978/bioe-591-genomics/course-materials/data/selection/metadata.csv")
x <- read.pcadapt("/home/f85j978/bioe-591-genomics/course-materials/data/selection/salvelinus.vcf", type = "vcf")
```

# Create Plot for PCA selection of K

```{r}
pca <- pcadapt(x, K = 20)
plot(pca, option = "screeplot")
```

# Setting K = 4 for outlier detection

```{r}
pca_4 <- pcadapt(x, K = 4)
```

# Identifying outliers and printing outlier and neutral loci totals

```{r}
alpha <- 0.05
padj <- p.adjust(pca_4$pvalues, method = "bonferroni")
outlier_idx <- which(padj < alpha)
neutral_idx <- which(padj >= alpha)

cat("Outlier loci:", length(outlier_idx), "\n")
cat("Neutral loci:", length(neutral_idx), "\n")
```

# Plot outliers relative to their order

```{r}
plot(pca_4)
```

# Load VCF and subset it with index of outliers and neutral loci

```{r}
vcf <- read.vcfR("/home/f85j978/bioe-591-genomics/course-materials/data/selection/salvelinus.vcf", verbose = FALSE)
vcf_neutral <- vcf[neutral_idx, ]
vcf_outlier <- vcf[outlier_idx, ]
```

# Converting SNPs to genid objects

```{r}
dna_neutral <- vcfR2DNAbin(vcf_neutral, unphased_as_NA = FALSE, consensus = TRUE, extract.haps = FALSE)
gi_neutral <- DNAbin2genind(dna_neutral)
dna_outlier <- vcfR2DNAbin(vcf_outlier, unphased_as_NA = FALSE, consensus = TRUE, extract.haps = FALSE)
gi_outlier <- DNAbin2genind(dna_outlier)
```

# Scale genotypes

```{r}
gi_neutral_scaled <- scaleGen(gi_neutral, NA.method = "mean", scale = FALSE)
pca_neutral <- prcomp(gi_neutral_scaled, center = FALSE, scale. = FALSE)
gi_outlier_scaled <- scaleGen(gi_outlier, NA.method = "mean", scale = FALSE)
pca_outlier <- prcomp(gi_outlier_scaled, center = FALSE, scale. = FALSE)
```

# Transform our PCA objects into dataframes by extracting the first two PC axes from each => making rownames into a column labeled sample => joining these to our metadata object (meta) by aligning sample names

```{r}
# turn lists into dataframes
neutral_df <- data.frame(pca_neutral$x[, 1:2]) %>%
  mutate(sample = rownames(.))
outlier_df <- data.frame(pca_outlier$x[, 1:2]) %>%
  mutate(sample = rownames(.))

# join metadata
neutral_df <- left_join(neutral_df, meta, by = c("sample" = "Sample.ID"))
outlier_df <- left_join(outlier_df, meta, by = c("sample" = "Sample.ID"))

# shared axis limits
xlims <- range(c(outlier_df$PC1, neutral_df$PC1), na.rm = TRUE)
ylims <- range(c(outlier_df$PC2, neutral_df$PC2), na.rm = TRUE)

# percent variance explained
neutral_var <- 100 * summary(pca_neutral)$importance[2, ]
outlier_var <- 100 * summary(pca_outlier)$importance[2, ]
```

# Compare PCA plots for outlier and neutral loci

```{r}
p1 <- ggplot(neutral_df, aes(x = PC1, y = PC2, color=Site.Latitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = xlims, ylim = ylims) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(
    title = "PCA: neutral loci",
    x = paste0("PC1 (", round(neutral_var[1], 1), "%)"),
    y = paste0("PC2 (", round(neutral_var[2], 1), "%)")
  ) 

p2 <- ggplot(outlier_df, aes(x = PC1, y = PC2, color=Site.Latitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = xlims, ylim = ylims) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(
    title = "PCA: outlier loci only",
    x = paste0("PC1 (", round(outlier_var[1], 1), "%)"),
    y = paste0("PC2 (", round(outlier_var[2], 1), "%)")
  )

p1 + p2
```

-----------------------------------------------------------------------------

Homework

1. Add commented text or a Markdown chunk below the PCA plot comparison with your interpretation of the differences between results from neutral loci and outlier loci.

**Both the netural and outlier loci show clear separation of the southernmost cluster of genetic samples. In both graphs it is apparent that PC1 separates the southermost cluster; however, the separation between the broader distribution of sites dissappears when you limit it to only include outlier loci.**

2. Plot both PC1 and PC2 for both datasets against latitude. In a comment or Markdown chunk, describe your results.

```{r}
#PC1 Neutral Loci and Latitude
pc1_lat_n <- ggplot(neutral_df, aes(x = Site.Latitude, y = PC1, color=Site.Latitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(neutral_df$Site.Latitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC1 and Latitude - Neutral Loci",
    x = "Site Latitude",
    y = paste0("PC1 (", round(neutral_var[1], 1), "%)")
  )
  
#PC1 Outlier Loci and Latitude
pc1_lat_o <- ggplot(outlier_df, aes(x = Site.Latitude, y = PC1, color=Site.Latitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(outlier_df$Site.Latitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC1 and Latitude - Outlier Loci",
    x = "Site Latitude",
    y = paste0("PC1 (", round(outlier_var[1], 1), "%)")
  ) 

pc1_lat_n + pc1_lat_o
```

```{r}
#PC2 Neutral Loci and Latitude
pc2_lat_n <- ggplot(neutral_df, aes(x = Site.Latitude, y = PC2, color=Site.Latitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(neutral_df$Site.Latitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC2 and Latitude - Neutral Loci",
    x = "Site Latitude",
    y = paste0("PC2 (", round(neutral_var[1], 1), "%)")
  )
  
#PC2 Outlier Loci and Latitude
pc2_lat_o <- ggplot(outlier_df, aes(x = Site.Latitude, y = PC2, color=Site.Latitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(outlier_df$Site.Latitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC2 and Latitude - Outlier Loci",
    x = "Site Latitude",
    y = paste0("PC2 (", round(outlier_var[1], 1), "%)")
  ) 

pc2_lat_n + pc2_lat_o
  
pc1_lat_n + pc1_lat_o + pc2_lat_n + pc2_lat_o 
```

**The neutral loci and latitude plots, lack clustering, and follow an isolation by distance pattern with samples in close proximity having similar PC1/PC2 values. However, when we consider the outlier loci plots there is clear separation of the southernmost populations. The southernmost population is best separated by the PC1 axis. PC2 separates another southern cluster from the rest of the distribution.**

3. Plot PC1 and PC2 for both datasets against longitude, again adding text to describe your results.

```{r}
#PC1 Neutral Loci and Longitude
pc1_long_n <- ggplot(neutral_df, aes(x = Site.Longitude, y = PC1, color=Site.Longitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(neutral_df$Site.Longitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC1 and Longitude - Neutral Loci",
    x = "Site Longitude",
    y = paste0("PC1 (", round(neutral_var[1], 1), "%)")
  )
  
#PC1 Outlier Loci and Longitude
pc1_long_o <- ggplot(outlier_df, aes(x = Site.Longitude, y = PC1, color=Site.Longitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(outlier_df$Site.Longitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC1 and Longitude - Outlier Loci",
    x = "Site Longitude",
    y = paste0("PC1 (", round(outlier_var[1], 1), "%)")
  ) 

pc1_long_n + pc1_long_o
```

```{r}
#PC2 Neutral Loci and Longitude
pc2_long_n <- ggplot(neutral_df, aes(x = Site.Longitude, y = PC2, color=Site.Longitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(neutral_df$Site.Longitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC2 and Longitude - Neutral Loci",
    x = "Site Longitude",
    y = paste0("PC2 (", round(neutral_var[1], 1), "%)")
  )
  
#PC2 Outlier Loci and Longitude
pc2_long_o <- ggplot(outlier_df, aes(x = Site.Longitude, y = PC2, color=Site.Longitude)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis() +
  coord_cartesian(xlim = range(outlier_df$Site.Longitude, na.rm = TRUE)) +
  theme_classic() +
  theme(legend.position = "bottom") #+
  labs(
    title = "PC2 and Longitude - Outlier Loci",
    x = "Site Longitude",
    y = paste0("PC2 (", round(outlier_var[1], 1), "%)")
  ) 

pc2_long_n + pc2_long_o

pc1_long_n + pc1_long_o + pc2_long_n + pc2_long_o
```

**Given the climatic gradient we are concerned with is best represented by a latitude, it is not as meaningful to plot the loci against latitude. There appears to be separation of a cluster near -80 degrees latitude, which likely corresponds to the southernmost populations.**